---
title: Tidy Evaluation in R (Part 3)
author: Mehmet Oner Yalcin
date: '2019-06-02'
slug: tidy-eval-r-part3
categories:
  - R
tags:
  - rlang
  - tidyeval
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---

If you have some experience wriitng R functions, there is possibility you may have hit a problem called unintended data masking. This can introduce subtle but hard to find bugs. We'll have a look at it in this post as well as advanced use of `!!` operator. Namely bypassing the data mask. 

As always it's best understood via an example. Below `multiply_height` is a very simple function, taking a dataframe named `data`, and (assuming there is `height`) multiplying `height` by `amount` and transmuting.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(rlang)
```

```{r}
multiply_height <- function(data, amount) {
  transmute(data, height*amount)
}

multiply_height(starwars, 10)
```


if you are creating a fuction for the package you have no idea what's going to be in the `data` and `data` is used for data mask. Any columns in the data will be masking the objects used in the arguments in the `transmute` call. Let's shadow the `amount` argument by creating an `amount` column.


```{r}
starwars %>% 
  mutate(amount = 0) %>% 
  multiply_height(10)
```


As seen above `amount` column in the data has higher priority over thge argument `amount` in the `transmute` call. Therefore we introduced a bug to our function. 

Unquote operator `!!` is useful at this point. Let's define our `multiply_height` function once again and unqoute `amount` argument

```{r}
multiply_height <- function(data, amount) {
  transmute(data, height * !!amount)
}

starwars %>% 
  mutate(amount = 0) %>% 
  multiply_height(10)
```

Now `amount` column doesn't impact the `argument` inside our function, our subtle but annoying bug is resolved. 

Hope you found it useful!

