---
title: Tidy Evaluation in R (Part 1)
author: Mehmet Oner Yalcin
date: '2019-05-31'
slug: tidy-eval-r
categories:
  - R
tags:
  - R
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---

This post is pretty much a transcript of [Lionel Henry](https://twitter.com/@_lionelhenry)'s [webinar](https://resources.rstudio.com/webinars/tidyeval) on rstudio. Therefore I recommend you to [watch it](https://resources.rstudio.com/webinars/tidyeval) if you prefer video rather than text. 

`rlang` library provides a set of functions to do some magic, it's best explained on examples:

Dataset `starwars` has a two particular columns `height` and `mass`

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(rlang)

starwars %>%  head()
```

if we want to find body/mass ratio then we can use baseR function `quote` to create a kind of symbolic formula 

```{r}
sym_formula <- quote(height/mass)
sym_formula
```
Note that above we didn't input `height` or `mass` as strings. 

Using baseR function `eval` we can run the symbolic expression on `starwars` dataset 

```{r}
eval(sym_formula, starwars)
```

Note that the expression above `eval(sym_formula, starwars)` equals to `starwars$height / starwars$mass`

Let's define a function called `cement`, I won't explain the particular details inside at the moment but it basically takes unquoted names and pastes them together with space:

```{r}
cement <- function(...) {
  vars <- rlang::ensyms(...)
  vars <- purrr::map(vars, as.character)
  paste(vars, collapse = " ")
}

cement(Oner, Yalcin)
```

The issue is that `cement` treats all unqoted names as constants, for example I'll add a new variable called `firstname`

```{r}
firstname <- "Mehmet"

cement(firstname, Oner, Yalcin)
```

In order to get `Mehmet Oner Yalcin` and evaluate `firstname` as variable we need to use `!!`. This is called **unquoting**

```{r}
cement(!!firstname, Oner, Yalcin)
```

Quasiquotation is the combination of quoting an expression while allowing immediate evaluation (unquoting) of part of that expression. To demystify this sentence we will show an example using `rlang::qq_show` 

```{r}
firstname 

qq_show(!!firstname)
```

above these two expressions are not equal, while `firstname` is a variable and holding string `Mehmet`, `qq_show` doesn't return anything but has a side effect, prints out `Mehmet` to console

```{r}
x <- qq_show(!!firstname)

print(x)
```

We can use `qq_show` to see the evaluated result of an quoted expression

```{r}
qq_show(
  cement(!!firstname, Oner, Yalcin)
)
```

### First Example: Working with Column names

*Goal: Compute average on different column names*

```{r}
starwars %>% 
  summarise(avg = mean(height, na.rm = TRUE))

cols <- c("height", "mass")
```

**Naive Attempt with `!!`**

```{r error=FALSE}
col <- cols[[1]]

starwars %>% 
  summarise(avg = mean(!!col, na.rm = TRUE))

```


Let's see how `!!col` was evaluated

```{r}
qq_show(
  starwars %>% 
    summarise(avg = mean(!!col, na.rm = TRUE))
)
```

as seen in the `qq_show` output above `!!col` argument evaluated as string `"height"`. This means we are trying to find the mean of string `"height"`, which is not defined as it is not numeric or logical

**Creating Symbols**

We'll introduce another function called `rlang::sym`, this function take string as input and turn it into symbol.

```{r}
sym("Mehmet")
```

Looks better, let's try again

```{r}
col <- sym(cols[[1]])

qq_show(
  starwars %>% 
    summarise(avg = mean(!!col, na.rm = TRUE))
)

```

yes, we are closer, at least now `height` is not a string. Let's run it without `qq_show`

```{r}
starwars %>% 
  summarise(avg = mean(height, na.rm = TRUE))
```

### Using `!!` in a loop

```{r}
dfs <- vector("list", length(cols))
for (i in seq_along(cols)) {
  col <- rlang::sym(cols[[i]])
  dfs[[i]] <- starwars %>% 
    summarise(avg = mean(!!col, na.rm = TRUE))
}

dfs
```

Ok it works but we don't fancy `for` loops in a functional programming language like R. Let's do it properly, using functions


```{r}

summarize_avg <- function(data, col) {
  col <- sym(col)
  data %>% 
    summarise(avg = mean(!!col, na.rm = TRUE))
}

summarize_avg(starwars, "height")
```

now we can use `purrr::map`

```{r}
map(cols, summarize_avg, data = starwars)
```


It works as expected, however it's not the prettiest. The problem with `summarize_avg` is that it takes the column name as string not symbol, let's say we use our function in an analysis pipeline like the example below:

```{r}
starwars %>% 
  group_by(species) %>% 
  summarize_avg("height")
```

It works but this is not coherent with dplyr interface, if we ommit the double quotes and feed `height` as an object then we get error

```{r error=TRUE}
starwars %>% 
  group_by(species) %>% 
  summarize_avg(height)
```